---
title: "Assignment 3"
author: "IM 532 3.0 Applied Time Series Forecasting"
date: "14/06/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1

Data set `fma::books` contains the daily sales of paperback and hardcover books at the same store.

```{r, comment=NA, message=FALSE}
library(fma)
books
```

The task is to forecast the next four days’ sales for paperback and hardcover books.

i) Plot the data and describe the main features of the series.

ii) Use an ETS(A,N,N) model to forecast each series, and plot the forecasts.

iii) Compute the RMSE values for the training data in each case.

## Question 2

For this exercise use data set `expsmooth::visitors`, the monthly Australian short-term overseas visitors data, May 1985–April 2005

```{r, comment=NA, message=FALSE}
library(expsmooth)
visitors
```

i. Make a time plot of your data and describe the main features of the series.

ii. Create a training set that withholds the last two years of available data.

iii. Use ARIMA, snaive, random walk, random walk with drift, average method on the time series. Which method gives the best forecasts? Give reasons for your answer.

**Due date: ** 20 June 2020


