---
title: "IM532 3.0 Applied Time Series Forecasting"
subtitle: "MSc in Industrial Mathematics"
author: "Thiyanga Talagala"
date: ""
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: 
      - default
      - default-fonts
      - duke-blue
      - hygge-duke
      - libs/cc-fonts.css
      - libs/figure-captions.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

## Time series decomposition

### Recall: Time series patterns

**Trend** ($T_t$)

Long-term increase or decrease in the data.


**Seasonal** ($S_t$)

A seasonal pattern exists when a series is influenced by seasonal factors (e.g., the quarter of the year, the month, or day of the week). Seasonality is always of a fixed and known period. Hence, seasonal time series are sometimes called periodic time series.

Period is unchanging and associated with some aspect of the calendar.

**Cyclic** 

A cyclic pattern exists when data exhibit rises and falls that are not of fixed period. The duration of these fluctuations is usually of at least 2 years.


<!--Not too much into theory. -->

---

# Time series decomposition

$$y_t = f(S_t, T_t, R_t)$$

$y_t$: data at period $t$.

$T_t$: trend-cycle component at period $t$.

$S_t$: seasonal component at period $t$.

$R_t$: remainder component at period $t$.

## Additive decomposition

$$y_t=S_t + T_t + R_t$$
## Multiplicative decomposition

$$y_t=S_t \times T_t \times R_t$$

To transform from multiplicative to additive, you can use a Box-Cox transformation (particularly, **log** transformation).

<!--Additive model appropriate if magnitude of **seasonal** fluctuations does not vary with level.-->

<!-- If **seasonal** are proportional to level of a series then multiplicative model appropriate.-->

<!-- Multiplicative decomposition more prevalent with economics models.-->

---
## Multiplicative vs. Additive

.pull-left[


]

.pull-right[


]

---

<!--Variation of seasonality change over time.-->

## `AirPassengers`


```{r, comment=NA, message=FALSE, warning=FALSE, echo=FALSE}
library(forecast)
library(fpp2)
fit <- stl(AirPassengers, s.window = 12)
autoplot(fit)
```

---
