---
title: "IM532 3.0 Applied Time Series Forecasting"
subtitle: "Dynamic regression models"
author: "Dr. Thiyanga Talagala"
date: "21 June 2020"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: 
      - default
      - default-fonts
      - duke-blue
      - hygge-duke
      - libs/cc-fonts.css
      - libs/figure-captions.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

# Introduction

- ARIMA models

- ETS models

- Dynamic regression models: How to include external variables

## Regression model

$$y_t=\beta_0 + \beta_1x_{1,t}+...+\beta_kx_{k,t}+\epsilon_t$$

## Allow the errors from a regression to contain autocorrelation.

$$y_t=\beta_0 + \beta_1x_{1,t}+...+\beta_kx_{k,t}+\eta_t$$
$$(1-\phi_1 B)(1-B)\eta_t=(1+\theta_1B)\epsilon_t$$

where $\epsilon_t$ is a white noise series.

$\eta_t$ follows an $ARIMA(1, 1, 1)$

---

# Stationary variables vs Nonstationary Variables

- If all of the variables in the model are stationary, then we only need to consider ARMA errors for the residuals

- Regression model with ARIMA errors is equivalent to a regression model in differences with ARMA errors.

$$y'_t=\beta_0 + \beta_1x'_{1,t}+...+\beta_kx'_{k,t}+\eta'_t$$

$$(1-\phi_1 B)(1-B)\eta'_t=(1+\theta_1B)\epsilon_t$$

$$y'_t=y_t - y_{t-1}$$

$$x'_{t,i}=x_{t, i} - x_{t-1, i}$$

$$\eta'_{t}=\eta_{t} - \eta_{t-1}$$

---

# Regression  with ARIMA errors in R

To fit

$$y'_t = \beta_1x'_t+\eta'_t,$$

where $\eta'_t=\phi_1\eta'_{t-1}+\epsilon_t$ is an $AR(1)$ error.

This is equivalent to 

$$y_t=\beta_0 + \beta_1x_t + \eta_t$$
where $\eta_t$ is an $ARIMA(1, 1, 0)$ error.

Constant term disappears due to the differencing.
```r
fit <- Arima(y, xreg=x, order=c(1,1,0))
```
---
# Using automated ARIMA algorithm

```{r, comment=NA, message=FALSE, warning=FALSE}
library(forecast)
library(fpp2)
head(uschange)

```

---
# Using automated ARIMA algorithm

```{r, comment=NA, message=FALSE, warning=FALSE, fig.height=5}
autoplot(uschange[,1:2], facets=TRUE) +
xlab("Year") + ylab("") +
ggtitle("Quarterly changes in US consumption
and personal income")
```

---
# Using automated ARIMA algorithm

```{r, comment=NA, message=FALSE, warning=FALSE}
fit <- auto.arima(uschange[,"Consumption"],
xreg=uschange[,"Income"])
fit
```

The fitted model is

$\hat{y}_t=0.599 + 0.203x_t + \eta_t$

$\eta_t=0.692\eta_{t-1}+\epsilon_t-0.576\epsilon_{t-1}+0.198\epsilon_{t-2}$

$\epsilon_t \sim NID(0, 0.322)$


---

# plot $\eta_t$ and $\epsilon_t$

```{r, comment=NA, message=FALSE, warning=FALSE, fig.height=4}
library(magrittr)
cbind("Regression Errors" = residuals(fit, type="regression"),
"ARIMA errors" = residuals(fit, type="innovation")) %>%
autoplot(facets=TRUE)
```

---

```{r, fig.height=4, comment=NA, message=FALSE, warning=FALSE, fig.height=4}
checkresiduals(fit)
```

---

# Calculate forecasts

1. We first need to forecast predictors

```{r, comment=NA, message=FALSE, warning=FALSE, fig.height=4}
fcast <- forecast(fit, xreg=rep(mean(uschange[,2]),8))
autoplot(fcast) + xlab("Year") + ylab("Percentage change")
```

---

# Forecasting electricity demand

```{r, comment=NA, message=FALSE, warning=FALSE, fig.height=4}
xreg <- cbind(MaxTemp = elecdaily[, "Temperature"],
MaxTempSq = elecdaily[, "Temperature"]^2,
Workday = elecdaily[, "WorkDay"])
fit <- auto.arima(elecdaily[, "Demand"], xreg = xreg)
fit
```

---

# Forecasting electricity demand (cont.)

```{r, comment=NA, message=FALSE, warning=FALSE, fig.height=4}
checkresiduals(fit)
```

---
# Lagged predictors

$$y_t = \beta_0 + \gamma_0x_t + \gamma_1x_{t-1}+...+\gamma_kx_{t-k}+\eta_t$$

where $\eta_t$ is an $ARIMA$ process. 

## How to select $k$ ? 

Use AICc along with the values of $p$ and $q$.

---

```{r, comment=NA, message=FALSE, warning=FALSE}
autoplot(insurance, facets=TRUE) +
xlab("Year") + ylab("") +
ggtitle("Insurance advertising and quotations")
```

---

## Lagged predictors

```{r, comment=NA, message=FALSE, warning=FALSE}
Advert <- cbind(
AdLag0 = insurance[,"TV.advert"],
AdLag1 = stats::lag(insurance[,"TV.advert"],-1),
AdLag2 = stats::lag(insurance[,"TV.advert"],-2),
AdLag3 = stats::lag(insurance[,"TV.advert"],-3)) %>%
head(NROW(insurance))

head(Advert)
```

---
## Lagged predictors (cont.)

```{r, comment=NA, message=FALSE, warning=FALSE}
fit1 <- auto.arima(insurance[4:40,1], xreg=Advert[4:40,1],
stationary=TRUE)
fit1[["aicc"]]
fit2 <- auto.arima(insurance[4:40,1], xreg=Advert[4:40,1:2],
stationary=TRUE)
fit2[["aicc"]]
fit3 <- auto.arima(insurance[4:40,1], xreg=Advert[4:40,1:3],
stationary=TRUE)
fit3[["aicc"]]
fit4 <- auto.arima(insurance[4:40,1], xreg=Advert[4:40,1:4],
stationary=TRUE)
fit4[["aicc"]]
```

---

## Lagged predictors (cont.)

Re-estimate the model

```{r, comment=NA, message=FALSE, warning=FALSE}
fit <- auto.arima(insurance[,1], xreg=Advert[,1:2],
stationary=TRUE); fit
```

The fitted model is

$$y_t = 2.039 + 1.25x_t + 0.162x_{t-1}+\eta_t$$
$$\eta_t=1.412\eta_{t-1}-0.932\eta_{t-2}+0.359\eta_{t-3}+\epsilon_t$$

$y_t$ - the number of quotations provided in month $t$

$x_t$ - the advertising expenditure in month $t$

$\epsilon_t$ is white noise.

---

# Generate forecasts

```{r, comment=NA, message=FALSE, warning=FALSE, fig.height=4}
fc8 <- forecast(fit, h=20,
xreg=cbind(AdLag0 = rep(8,20),
AdLag1 = c(Advert[40,1], rep(8,19))))
autoplot(fc8) + ylab("Quotes") +
ggtitle("Forecast quotes with future advertising set to 8")

```

---

```{r, comment=NA, message=FALSE, warning=FALSE, fig.height=4}
checkresiduals(fc8)
```

---

# Other approaches

- Neural network models

- Machine learning approaches

- TBATS models

- Forecasting with decomposition

- Vector autoregressions