---
title: "IM532 3.0 Applied Time Series Forecasting"
subtitle: "MSc in Industrial Mathematics"
author: "Dr. Thiyanga Talagala"
date: "1 March 2020"
output:
  xaringan::moon_reader:
    chakra: libs/remark-latest.min.js
    lib_dir: libs
    css: 
      - default
      - default-fonts
      - duke-blue
      - hygge-duke
      - libs/cc-fonts.css
      - libs/figure-captions.css
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---
```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```


# Time series

- A time series is a sequence of observations taken sequentially in time. 


---

## Deterministic  vs Non-deterministic time series

.pull-left[
- **Deterministic time series:** future values can be exactly determined by using some mathematical function.



```{r, echo=FALSE, comment=NA, message=FALSE}
library(ggplot2)
library(forecast)
x <- 0:5000
y <- cos(2*3.14*x)
y <- ts(y, frequency=1)
autoplot(y)

```


$$y_t = cos(2\pi t)$$

]

.pull-right[
- **Non-deterministic time series:** future values can be determined only in terms of a probability distribution.



```{r, echo=FALSE, comment=NA, message=FALSE}
library(mozzie)
y <- ts(mozzie$Colombo)
#y <- ts(mozzie$Colombo, frequency=52, start=c(2008, 51))
autoplot(y)+ylab("Number of dengue cases in Colombo")

```

]

---

## Stochastic processes

"A statistical phenomenon that evolves in time according to probabilistic laws is called a stochastic process." (Box, George EP, et al. Time series analysis: forecasting and control.)
          
##  Non-deterministic time series or statistical time series

A sample realization from an infinite population of time series that could have been generated by a stochastic process. 

```{r, echo=FALSE, include=TRUE, results="asis"}
library(knitr)
library(magrittr)
library(kableExtra)
mathy.df <- data.frame(a = c("Sequence of random Variables", "Realized values of the random Variable", "Examples of realizations", "", "", ""), 
b = c('$\\{X_1, X_2...X_t...X_N\\}$', "$\\{x_1, x_2,...x_t...x_N\\}$", "{20, 40, 20..., 190}", "{15, 20, 100..., 490}", "{15, 44, 39, ..., 329}", "...")
)
colnames(mathy.df) <- c("Description", "Notation")
kable(mathy.df, format="html", align = "l", booktabs = T, escape = FALSE) %>% add_footnote("N - length of the time series, t - time index")
```

---

- 

## Time series to be analyzed


.pull-left[
```{r, comment=NA, echo=F}
y <- ts(mozzie$Colombo, frequency=52, start=c(2008, 51))
y
```

]

.pull-right[
```{r, comment=NA, echo=FALSE}
autoplot(y)+ylab("Number of dengue cases in Colombo")
```
]

- The observed time series or time series to be analyzed is **a particular realization of a stochastic process**.
---



---
class: center, middle


