---
title: "Tutorial 1"
author: "IM532 3.0 Applied Time Series Forecasting"
date: "07/04/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE, warning = FALSE)
```


## Data

Use data in `colmozzie` package in R. 

## Package installation

```{r, eval=FALSE}
install.packages("colmozzie")
```

```{r}
library(colmozzie)
head(colmozzie)
```

To view the description of variables type,

```{r, eval=FALSE}
?colmozzie
```

## Other packages

```{r, comment=NA, eval=FALSE}
install.packages("forecast")
library(forecast)
```



1. Use TEM: average temperature of the week (Celcius) in `colmozzie` data set. Split the data into two parts: i) training set (2009 to 2013 data) and ii) test set (2014 data).

```{r, comment=NA}
tem <- ts(colmozzie$TEM, start=c(2009, 1), frequency = 52)
tem.train <- window(tem, end=c(2013, 52))
tem.test <- window(tem, start=c(2014, 1))
```

2. The following code can be used to check whether time series has been split correctly.

```{r, comment=NA}
library(forecast)
autoplot(tem) +
autolayer(tem.train, series="Training") +
autolayer(tem.test, series="Test")
```

4. Calculate forecasts using `naive` method applied to `tem.train` .

```{r, comment=NA}
forecast.naive <- naive(tem.train, h=length(tem.test))
forecast.naive
```

5. Compare the accuracy on the results using the test set.

```{r, comment=NA}
accuracy(forecast.naive, tem.test)
```

6. Check the residuals. Do they resemble white noise?

```{r, comment=NA}
checkresiduals(forecast.naive)
```


7. Use other benchmark methods such as snaive, average method and random walk with drift methods to forecast the training set and compare the results on the test set. 

8. Which do you think is best to forecast average temperature in Colombo district? Give reasons for your answer.

9. Repeat the exercise for the other climate variable (TMAX, Tm, SLP, H, PP, VV, V, VM) in `colmozzie` dataset.

**Note:**

**The above exercise is prepared based on Hyndman, R. J., & Athanasopoulos, G. (2018). Forecasting: principles and practice. OTexts. adapting to colmozzie data set avilable on colmozzie package in R written by Thiyanga S. Talagala.**

