---
title: "Linear Time Series Analysis and Its Applications - Part 2" 
author:
- familyname: Talagala
  othernames: Dr Thiyanga
  address: University of Sri Jayewardenepura
  email: ttalagala@sjp.ac.lk
- familyname: 
  othernames: Note
  address: IM 532 3.0 Applied Time Series Forecasting, online teaching material - 1 during the COVID-19 outbreak. This is designed for self-study to support your learning.
abstract: ""
keywords: ""
wpnumber: no/yr
jelcodes: C10,C14,C22
blind: true
cover: true
toc: true
bibliography: references.bib
biblio-style: authoryear-comp
output:
  MonashEBSTemplates::workingpaper:
    fig_caption: yes
    fig_height: 5
    fig_width: 8
    includes:
      in_header: preamble.tex
    keep_tex: yes
    number_sections: yes
    citation_package: biblatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache=TRUE, messages=FALSE, warning=FALSE)
# Make sure you have the latest version of rmarkdown and bookdown
#devtools::install_github("rstudio/rmarkdown")
#devtools::install_github("rstudio/bookdown")
library(ggplot2)
```

# Introduction: Different Types of Non-Stationary Models


## Random walk 

$$Y_t = Y_{t-1} + \epsilon_t$$

### Random walk process: Mean

$$E(Y_t) = E(Y_{t-1} + \epsilon_t)$$ 

 - non-mean-reverting process:  move away from the mean either in a positive or negative direction.
 
### Random walk process: Variance
 
 - the variance evolves over time and goes to infinity as time goes to infinity;
 
### Simulation

```{r, comment=NA, messages=FALSE, eval=FALSE}
library(forecast)
y <- cumsum(rnorm(20))
y <- as.ts(y)
autoplot(y)+ggtitle("Random walk process")
```

## Random walk with drift


### Simulation


## Deterministic trend



### Simulation

```{r, message=FALSE, echo=TRUE}
library(forecast)
t <- 1:20
y <- 5 + 10*t + rnorm(20)
y <- as.ts(y)
autoplot(y)+ggtitle("Deterministic trend process")
```

## Random walk with drift and deterministic trend

# How to convert a nonstationary series to a stationary time series?

## Difference stationary

A random walk with or without a drift can be transformed to a stationary process by
differencing (subtracting $Y_{t-1}$ from $Y_t$, taking the difference $Y_t - Y_{t-1}$).

Let, $X_t$ be the difference series of a random walk process. Then,

$$X_t=Y_t - Y_{t-1} = \epsilon_t.$$

We can see $X_t$ is equal to a white noise process. 

Similarly, the difference series of a random walk with drift process is

$$X_t=Y_t - Y_{t-1} = \alpha + \epsilon_t,$$
where $\alpha$ is a constant. Further for 


## Trend stationary process


## Unit root tests

# Model Identification

